---
title: "Simulation more complex architectures in defined gene sets (vcf & bed input)"
author: "Marcin Kierczak"
date: "03/19/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#library('vcfR')
library('stringr')
library('gwasim')
library('reticulate')
```

## Synopsis
This document documents a bit more advanced example analyses of defined gene sets (bed file) in a whole genome (VCF file) using the `gwasim` package.
Package can be installed by typing: `devtools::install_github("NBISweden/a_johansson_2020")`. For further details visit [gwasim GitHub pages](https://github.com/NBISweden/a_johansson_2020).

## Setup
Before starting our simulation, we need to load the necessary packages (install them earlier if you do not already have them in your library). The easiest way is to:

* `install.packages('renv')`,
* download this Renv [lock file](https://github.com/NBISweden/a_johansson_2020/blob/master/renv.lock),
* `renv::restore()`

Next, load the packages:
```{r setup_noexec, eval=FALSE}
library('vcfR')
library('stringr')
library('data.table')
library('gwasim')
```

## Reading data

Because VCF files can be really large, we will have to emplot the windowing strategy, i.e. read a number of variants into memory, check if they are in any region of interest and discard the variants outside pre-defined regions.

### Read genic regions of interest from bed file

In this step, we will load gene regions of interest, in this case, promotor regions, and match them to the available genomic data. In this case, we will focus on only one chromosome, chr22. Data defining our regions is stored in a `.bed` file. We need to remember that bed files have their coordinates based at 0 while VCF files are 1-based (see, e.g. [details](https://www.biostars.org/p/84686/))
```{r read_bed}
bed_data <- read_delim("~/Projects/johansson/promoters.bed", delim = " ", col_names = c('chr', 'start', 'end', 'feature'))

# Convert to 1-based coords used by the VCF format, filter out unnecessary data
bed_data %<>% 
  mutate(start = start + 1) %>% 
  select(-feature) %>%
  filter(chr == "chr22")
```

## Read VCF in chunks
For the sake of learning how to use `gwasim` we will read first 500 rows (markers) from the `ALL.chr22.phase3_shapeit2_mvncall_integrated_v5a.20130502.genotypes.vcf.gz` file. This file contains 1000 genomes phase 3 data for human chromosome 22. The file can be downloaded from the 1000 Genomes data repository via FTP [link](ftp://ftp.1000genomes.ebi.ac.uk/vol1/ftp/release/20130502/).

> NOTE! For this part of the analysis, we need to setup Python whic we will use for improved efficiency of reading indexed VCF files in chunks.

```{r setup_python_noexec, eval = F}
reticulate::py_install('pandas', pip = T)
reticulate::py_install('pyvcf', pip = T)
```

```{r load_data, cache=TRUE}
path <- "/Users/kiero/Projects/johansson/ALL.chr22.phase3_shapeit2_mvncall_integrated_v5a.20130502.genotypes.vcf.gz"
reader <- reticulate::import("vcf_region_reader")
i <- 1
region <- paste0(bed_data[i,'chr'], ':', bed_data[i,'start'], '-', bed_data[i,'end']) %>%
  str_remove("chr")
data <- reader$vcf_region_reader(reticulate::r_to_py(path), region)
G <- data %>% 
  gwasim::fix_allele_encoding() %>% 
  gwasim::impute_G()
G <- as.matrix(apply(G, 2, as.integer))
if (!('gwasim' %in% class(G))) class(G) <- c(class(G), 'gwasim')
maf <- get_maf(x)
```

## Simulate effect of rare alleles for genotype
Now, it is time to do some basic simulations, i.e. to simulate phenotypic values based on the genotypes in our `G`. 
To this end, we will use `gwasim::get_effects`. In this simple example (we will show more complex scenarios later in this document), we would like to simulate the following genetic architecture affecting phenotype:

* there are two randomly selected loci in the region that affect the phenotype,
* loci are selected only from the rare alleles with $maf <= 0.01$,
* the effects of the two loci are drawn from Beta distribution with shape parameters set to .1 and .1,
* all effects are positive, i.e. both selected loci have an additive positive effect.

```{r simulate_effects}
effects <- gwasim::get_effects(maf=maf, thr=0.01, N=2, get_betas_fun = dbeta, get_betas_args = list(shape1 = .1, shape2 = .1), rare=T, frac_negative=0, seed = 2)
effects
```
Here, we can see that, indeed, two markers have been selected: `rs587720402` and `rs587638893` and the effects at these loci are: $\beta_{rs587720402} = 13.66754$ and $\beta_{rs587638893} = 31.14350$ respectively. We are than ready to simulate our phenotype.

## Compute simulated phenotypes

First, we want some noise in our phenotype to simulate the effect of environment, other loci or simply the measurement error. Let us add some very tiny noise here sampled from $\mathcal{N}(0, 1)$.
```{r compute_phenos}
error_term <- rnorm(n = dim(G)[1], mean = 0, sd = 1)
# Check the error is as you want it
plot(error_term, pch=19, cex=.5, las=1)
hist(error_term)
```
Luckily, our error seems to be what we want -- it looks normally distributed around 0 and also it is pretty homogenous.
Having this last bit ready, we can simulate our phenotype according to the $\hat{y} = G \times \beta + \epsilon$:
```{r}
# Simulate phenotypes (here we do not want negative phenotype values so, we set them to 0)
y <- pmax(0, G[, effects$marker_idx] %*% effects$effects + error_term)
plot(y, 1:length(y), pch = 1 + rowSums(G[, effects$marker_idx]), xlab='individual', ylab='phenotype', las=1, cex.axis=.7, main = "Simulated phenotype per individual")
```

Indeed, we see there are a number of `rs587720402` heterozygotes with phenotype close to the effect of $13.67 + \epsilon$ and one heterozygote at `rs587638893` with phenotype close to the corresponding effect of $58.09 + \epsilon$.
We can check it matches the values in the $G$ matrix:
```{r check_in_G}
G[rowSums(G[, effects$marker_idx]) > 0, effects$marker_idx]
```

## Testing GWAS models.

Now, it'd be really interesting to see whether a tool for genome-wide association studies, e.g. GenABEL can single out our markers that were used for simulation. Let's try. Now things are becoming exciting and one can start asking different research questions. 

```{r data_to_gwaa, cache=TRUE}
library('GenABEL')
test_header <- tibble(
       chr = c(22),
       pos = as.numeric(coords[1:10]),
       snp = colnames(G),
       strand = c('u')
)
test_G <- as_tibble(t(G))
test_data <- cbind(test_header, test_G)
data <- gwasim::tibble_to_gwaa(test_data, sex=rep(2, times=dim(G)[1]), trait=y, progress = F)
qt <- GenABEL::qtscore(y~1, data)
plot(qt@annotation$Position,-log10(qt@results$P1df), pch=19, cex=.5)
best <- which.min(qt@results$P1df)
qt@results[best,]
```
What happened with our `rs587720402` marker, it did not pop up in the Manhattan plot. Apparently its effect got masked... Let us do a conditional analysis this time and see whether it does help:
```{r cond_analysis}
top_gt <- as.double(data@gtdata[,'rs587670191'])
qt_cond <- GenABEL::qtscore(y~top_gt, data)
plot(qt_cond@annotation$Position,-log10(qt_cond@results$P1df))
points(qt@annotation$Position,-log10(qt@results$P1df), pch=19, cex=.5)
best <- which.min(qt_cond@results$P1df)
qt_cond@results[best,]
```
Hurray! Our conditional analysis worked! On the plot above, you can see old, non-conditional results as black filled points while the results of the conditional analysis are in shown as hollow circles.
