---
title: "power_simuations"
author: "Marcin Kierczak"
date: "9/11/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r pkgs, echo = F, message=FALSE, warning=FALSE} 
library(SKAT)
library(tidyverse)
```

## Purpose

Here, I perform a number of power simulations using SKAT built-in functions. There is a number of parameters that change and each plot corresponds to such set of parameters. In particular, I look at the following:
 
 * cutoff value for *maf*. All loci with `maf < cutoff` will be considered causal,
 * a percentage of causal SNPs (among all SNPs with maf < cutoff),
 * `BetaType` -- type of effect sizes. 
 * `MaxBeta` -- max effect size (default = 1.6) 
 * `NegativePercent` -- percentage of negative coefficients.

## Effect sizes

First, let's look at the distribution of effect sizes: 
 
```{r effects}
maf_vector <- seq(0.001, 0.05, by = 0.001) # generate a vector of maf values to plot betas for
betas <- data.frame(maf = maf_vector, beta = rep(NA, times = length(maf_vector)))
betas[, 'beta'] <- SKAT:::Get_Beta(Type = 'Log', MAF = maf_vector, MaxValue = 1.6, Sign = 0)
ggplot(betas,  mapping = aes(x=maf, y=beta)) + geom_line() + geom_point()
```

## Tests
depending on the maf cutoff for rare alleles. We test the following cutoffs:

 * 0.01
 * 0.05
 * $\frac{1}{\sqrt{(2N)}}\ =\ 0.022$ (NSHPS size $N = 1069$ individuals)
 
## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
